Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-05 tex/main.tex
more

\newcommand{\con}[1]{\mathsf{#1}}

@@

@@

\newcommand{\sfuns}[1]{#1^\mathrm{f}}
\newcommand{\sfundefs}[1]{#1^\mathrm{df}}
\newcommand{\sfunndefs}[1]{#1^\mathrm{nf}}
\newcommand{\sortint}{\mathrm{Int}}

@@

\newcommand{\Bool}{\con{Bool}}
\newcommand{\ltrue}{\top}
\newcommand{\lfalse}{\bot}
\newcommand{\lite}{\con{ite}}

@@

\newcommand{\Sigmalia}{\Sigma_{LIA}}
\newcommand{\extendsig}[1]{\mathcalx{E}( #1 )}

@@

Unfortunately, these fragments cannot accommodate the vast majority of
quantified formulas that correspond to recursive function definitions.
In particular, the fragments in~\cite{GeDeM-CAV-09} are not applicable when the argument of a recursive function is used in various ways (e.g. within an arithmetic expression), 
while the fragment in~\cite{ReyEtAl-1-RR-13} is not applicable for functions over infinite domains such as the integers.
So, for instance, the quantified formula $\forall x\typ{\sortint}.\;\, \lite( p \leq 0, p( x ) \teq 0, p( x ) \teq 2 * p( x - 1 ) )$ 
is not in the essentially uninterpreted fragment (since $x$ occurs as a child of $-$),
nor does finite model finding apply (since $x$ is of type $\sortint$).

@@

For instance, the set $\{ \forallf{f} \vec x. f( x ) \teq 0 \}$ is admissible.
A set of admissible function definitions can include definitions of non-terminating functions, e.g.
$\{ \forallf{f} x^\sortint. f( x ) \teq f( x - 1 ) \}$ or even $\{ \forallf{f} x. f( x ) \teq f( x ) \}$ are admissible.
On the other hand, the set $\{ \forallf{f} x. f( x ) \teq f( x ) + 1 \}$ is not admissible, 
since the empty set of formulas is closed under function expansion with respect to this set,
and there is no model satisfying $\forallf{f} x. f( x ) \teq f( x ) + 1$.
Consider the set $\{ \forallf{f} x. f( x ) \teq f( x ), \forallf{g}. \lite( \neg f( x ) \teq 5, g( x ) \teq g( x ) + 1, g( x ) \teq f( x ) ) \}$.
While this set has a model where $f$ and $g$ are interpreted as the constant $5$, it is not admissible
since the formula $f( 0 ) \teq 4$ is closed under function expansion with respect to this set,
but there exists no interpretation satisfying both $f( 0 ) \teq 4$ and $\forallf{g}. \lite( \neg f( x ) \teq 5, g( x ) \teq g( x ) + 1, g( x ) \teq f( x ) )$.

@@

\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv_0( t\typ{\tau},\, p )$ $=$
 \begin{itemize}
   \item[] $\mathsf{if}$ $\tau = \Bool$ and $t = \boolop(t_1,\ldots,t_n)$ $\mathsf{then}$
    \begin{itemize}
      \item[] $\mathsf{let}$ $( t'_i,\, \chi_i ) = \conv_0( t_i,\, \pol( \boolop,\, i,\, p ) )$ $\mathsf{for}$ $i = 1, \ldots, n$ $\mathsf{in}$
      \item[] $\mathsf{let}$ $\chi = \chi_1 \mathrel\wedge \cdots \mathrel\wedge \chi_n$ $\mathsf{in}$
      \item[] $\mathsf{if}$ $p = \ptrue$ $\mathsf{then}$
      \begin{itemize}
        \item[] \return \bigtuple{$\boolop(t'_1, \ldots, t'_n) ) \wedge \chi,\, \true$}
      \end{itemize}
      \item[] $\mathsf{else}$ $\mathsf{if}$ $p = \pfalse$ $\mathsf{then}$
      \begin{itemize}
        \item[] \return \bigtuple{$\boolop(t'_1, \ldots, t'_n) \vee \neg \chi,\, \true$}
      \end{itemize}
      \item[] $\mathsf{else}$
      \begin{itemize}
        \item[] \return \bigtuple{$\boolop(t'_1, \ldots, t'_n),\, \chi$}
      \end{itemize}
    \end{itemize}
  \item[] $\mathsf{else}$ $\mathsf{if}$ $t = \forallf{\con{f}} \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \item[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$
      %\item[] $\mathsf{assert}( \chi = \true )$;
      \item[] \return \bigtuple{$\forall y\typ{\fargsort{\con{f}}}.\; u' [ \vecfarg{\con{f}}( y ) / \vec x ],\, \true$}
    \end{itemize}
  \item[] $\mathsf{else}$ $\mathsf{if}$ $t = \forall \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \item[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$
      \item[] \return \bigtuple{$\forall \vec x.\; u',\, \forall \vec x.\; \chi$}
    \end{itemize}
   \item[] $\mathsf{else}$
   \begin{itemize}
     \item[] \return \bigtuple{$t,\, \bigwedge\, \{ \exists z : \fargsort{\con{f}}.\; \vecfarg{\con{f}}( z ) \teq \vec t \mid \con{f}( \vec t ) \in \terms( t ),\, \con{f} \in \sfundefs{\Sigma} \}$}
   \end{itemize}
 \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv( \varphi )$ $=$
 \begin{itemize}
   \item[] $\mathsf{let}$ $( \psi,\, \_ ) = \conv_0( \varphi,\, \ptrue )$ $\mathsf{in}$
   %\item[] $\mathsf{assert}( D = \emptyset )$;
   \item[] \return $\psi$
  \end{itemize}
\end{enumerate}

@@

If $t$ is an application of an interpreted predicate $\boolop
\in \{{\neg}, {\land}, {\lor}, {\teq}, {\lite}\}$,
we recursively call $\conv$ on the arguments $t_i$ and polarity $\pol( \boolop,
i, p )$, where $\pol$ is defined as
\begin{equation*}
\pol( \boolop, i, p ) =
\begin{cases}
p & \text{if either $\boolop \in \{{\land}, {\lor}\}$ or $\boolop = \lite$ and $i \in \{2, 3\}$} \\[-\jot]
\pneg p & \text{if $\boolop = {\neg}$} \\[-\jot]
\pnone & \text{otherwise}
\end{cases}
\end{equation*}
The negation $\pneg p$ of a polarity $p$ is $\pfalse$ if $p$ is $\ptrue$, $\ptrue$
if $p$ is $\pfalse$, and $\pnone$ if $p$ is $\pnone$.

@@

\begin{example}
\label{ex:translation}
Consider the (combined) signature of linear arithmetic and uninterpreted functions $\Sigma = \Sigmalia \cup \Sigma_{\euf}$,
where $\ssorts{\Sigma_{\euf}} = \{ \sortint \}$ and 
$\sfuns{\Sigma_{\euf}} = \{ s\typ{ \sortint \rightarrow \sortint }, c\typ{ \sortint } \}$.
Consider the $\Sigma$-formula $\varphi$:
\begin{equation} \label{eq:ex-before}
\forall_{s} x\typ{\sortint}.\; \lite( x \leq 0, s(x) \teq 0, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{equation} 
%
Here, $s$ is a function that returns the sum of positive numbers between $0$
and its argument $x$. The formula $\varphi$ is in definitional form with
respect to functions $\sfundefs{\Sigma_u}$, and states that the sum of all
positive numbers between $0$ and $c$ is greater than $100$. This formula is
satisfiable with a model that interprets $c$ as an integer greater than or
equal to $14$. Due to the universally quantified formula in the left disjunct,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find
a model for this formula. The signature $\extendsig{\Sigma}$ includes the type
$\fargsort{s}$, and the uninterpreted function $\farg{s}$ of type $\fargsort{s}
\rightarrow \sortint$. The result of $\conv( \varphi, \ptrue )$ is the pair $(
\psi, \emptyset )$, where after simplification, $\psi$ is the
$\extendsig{\Sigma}$-formula
%
\begin{eqnarray} \label{eq:ex-after}
\begin{split}
\forall y\typ{\fargsort{s}}.\;\, \lite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( c ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c )
\end{eqnarray} 
%
The universal quantification in the resulting formula is over an uninterpreted
type $\fargsort{s}$, thus enabling finite model finding
techniques~\cite{ReyEtAl-1-RR-13} to search for a finite interpretation for
$\fargsort{s}$. Moreover, all occurrences of $y$ in this formula are beneath
applications of the uninterpreted function $\farg{s}$, implying that the
formula is in the essentially uninterpreted fragment of~\cite{GeDeM-CAV-09},
for which there exists a complete instantiation procedure. We observe that both
\cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}), and both
produce a model for formula~(\ref{eq:ex-after}) in $<0.1$ second. $\square$
\end{example}

>>>>

\newcommand{\con}[1]{\mathsf{#1}}
\let\const=\con

@@

\let\oldvee=\vee
\def\vee{\mathrel{\oldvee}}
\let\oldwedge=\wedge
\def\wedge{\mathrel{\oldwedge}}

@@

\newcommand{\sfuns}[1]{#1^\mathrm{f}}
\newcommand{\sfundefs}[1]{#1^\mathrm{df}}
\newcommand{\sfunndefs}[1]{#1^\mathrm{nf}}
\newcommand{\sortint}{\ty{Int}}

@@

\newcommand\ty[1]{\con{#1}}
\newcommand{\Bool}{\ty{Bool}}
\newcommand{\ltrue}{\top}
\newcommand{\lfalse}{\bot}
\newcommand{\lite}{\con{ite}}

@@

\newcommand{\Sigmalia}{\Sigma_{\mathcalx{LIA}}}
\newcommand{\extendsig}[1]{\mathcalx{E}( #1 )}

@@

Unfortunately, these fragments cannot accommodate the vast majority of
quantified formulas that correspond to recursive function definitions.
In particular, the fragments in~\cite{GeDeM-CAV-09} are not applicable when the argument of a recursive function is used in various ways (e.g. within an arithmetic expression), 
while the fragment in~\cite{ReyEtAl-1-RR-13} is not applicable for functions over infinite domains such as the integers.
So, for instance, the quantified formula $\forall x : {\sortint}.\;\, \lite( p \leq 0, p( x ) \teq 0, p( x ) \teq 2 * p( x - 1 ) )$ 
is not in the essentially uninterpreted fragment (since $x$ occurs as a child of $-$),
nor does finite model finding apply (since $x$ is of type $\sortint$).

@@

For instance, the set $\{ \forallf{f} \vec x. f( x ) \teq 0 \}$ is admissible.
A set of admissible function definitions can include definitions of non-terminating functions, e.g.
$\{ \forallf{f} x : \sortint.\;\, f( x ) \teq f( x - 1 ) \}$ or even $\{ \forallf{f} x. f( x ) \teq f( x ) \}$ are admissible.
On the other hand, the set $\{ \forallf{f} x. f( x ) \teq f( x ) + 1 \}$ is not admissible, 
since the empty set of formulas is closed under function expansion with respect to this set,
and there is no model satisfying $\forallf{f} x. f( x ) \teq f( x ) + 1$.
Consider the set $\{ \forallf{f} x. f( x ) \teq f( x ), \forallf{g}. \lite( \neg f( x ) \teq 5, g( x ) \teq g( x ) + 1, g( x ) \teq f( x ) ) \}$.
While this set has a model where $f$ and $g$ are interpreted as the constant $5$, it is not admissible
since the formula $f( 0 ) \teq 4$ is closed under function expansion with respect to this set,
but there exists no interpretation satisfying both $f( 0 ) \teq 4$ and $\forallf{g}. \lite( \neg f( x ) \teq 5, g( x ) \teq g( x ) + 1, g( x ) \teq f( x ) )$.

@@

\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv_0( t\typ{\tau},\, p )$ $=$
 \begin{itemize}
   \item[] $\mathsf{if}$ $\tau = \Bool$ and $t = \boolop(t_1,\ldots,t_n)$ $\mathsf{then}$
    \begin{itemize}
      \item[] $\mathsf{let}$ $( t'_i,\, \chi_i ) = \conv_0( t_i,\, \pol( \boolop,\, i,\, p ) )$ $\mathsf{for}$ $i = 1, \ldots, n$ $\mathsf{in}$
      \item[] $\mathsf{let}$ $\chi = \chi_1 \wedge \cdots \wedge \chi_n$ $\mathsf{in}$
      \item[] $\mathsf{if}$ $p = \ptrue$ $\mathsf{then}$
      \begin{itemize}
        \item[] \return \bigtuple{$\boolop(t'_1, \ldots, t'_n) ) \wedge \chi,\, \true$}
      \end{itemize}
      \item[] $\mathsf{else}$ $\mathsf{if}$ $p = \pfalse$ $\mathsf{then}$
      \begin{itemize}
        \item[] \return \bigtuple{$\boolop(t'_1, \ldots, t'_n) \vee \neg \chi,\, \true$}
      \end{itemize}
      \item[] $\mathsf{else}$
      \begin{itemize}
        \item[] \return \bigtuple{$\boolop(t'_1, \ldots, t'_n),\, \chi$}
      \end{itemize}
    \end{itemize}
  \item[] $\mathsf{else}$ $\mathsf{if}$ $t = \forallf{\con{f}} \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \item[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$
      %\item[] $\mathsf{assert}( \chi = \true )$;
      \item[] \return \bigtuple{$\forall y\typ{\fargsort{\con{f}}}.\; u' [ \vecfarg{\con{f}}( y ) / \vec x ],\, \true$}
    \end{itemize}
  \item[] $\mathsf{else}$ $\mathsf{if}$ $t = \forall \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \item[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$
      \item[] \return \bigtuple{$\forall \vec x.\; u',\, \forall \vec x.\; \chi$}
    \end{itemize}
   \item[] $\mathsf{else}$
   \begin{itemize}
     \item[] \return \bigtuple{$t,\, \bigwedge\, \{ \exists z : \fargsort{\con{f}}.\; \vecfarg{\con{f}}( z ) \teq \vec t \mid \con{f}( \vec t ) \in \terms( t ),\, \con{f} \in \sfundefs{\Sigma} \}$}
   \end{itemize}
 \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv( \varphi )$ $=$
 \begin{itemize}
   \item[] $\mathsf{let}$ $( \psi,\, \_ ) = \conv_0( \varphi,\, \ptrue )$ $\mathsf{in}$
   %\item[] $\mathsf{assert}( D = \emptyset )$;
   \item[] \return $\psi$
  \end{itemize}
\end{enumerate}

@@

If $t$ is an application of an interpreted predicate $\boolop
\in \{{\neg}, {\wedge}, {\vee}, {\teq}, {\lite}\}$,
we recursively call $\conv$ on the arguments $t_i$ and polarity $\pol( \boolop,
i, p )$, where $\pol$ is defined as
\begin{equation*}
\pol( \boolop, i, p ) =
\begin{cases}
p & \text{if either $\boolop \in \{{\wedge}, {\vee}\}$ or $\boolop = \lite$ and $i \in \{2, 3\}$} \\[-\jot]
\pneg p & \text{if $\boolop = {\neg}$} \\[-\jot]
\pnone & \text{otherwise}
\end{cases}
\end{equation*}
The negation $\pneg p$ of a polarity $p$ is $\pfalse$ if $p$ is $\ptrue$, $\ptrue$
if $p$ is $\pfalse$, and $\pnone$ if $p$ is $\pnone$.

@@

\begin{example}
\label{ex:translation}
Consider the combined signature of linear arithmetic and uninterpreted
functions $\Sigma = \Sigmalia \cup \Sigma_{\euf}$, where
$\ssorts{\Sigma_{\euf}} = \{ \sortint \}$ and
$\sfuns{\Sigma_{\euf}} = \{ \con{s} : { \sortint \rightarrow \sortint },\, \const{c} : { \sortint } \}$.
Let $\varphi$ be the $\Sigma$-formula
\begin{equation} \label{eq:ex-before}
\forallf{\con{s}} x : {\sortint}.\;\, \lite( x \leq 0,\, s(x) \teq 0,\, s( x ) \teq x + s( x-1 ) ) \wedge s( c ) > 100
\end{equation} 
%
Here, $s$ is a function that returns the sum of positive numbers between $0$
and its argument $x$. The formula $\varphi$ is in definitional form with
respect to functions $\sfundefs{\Sigma_u}$, and states that the sum of all
positive numbers between $0$ and $c$ is greater than $100$. This formula is
satisfiable with a model that interprets $c$ as an integer greater than or
equal to $14$. Due to the universally quantified formula in the left disjunct,
current SMT techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are unable to find
a model for this formula. The signature $\extendsig{\Sigma}$ includes the type
$\fargsort{s}$, and the uninterpreted function $\farg{s}$ of type $\fargsort{s}
\rightarrow \sortint$. The result of $\conv( \varphi, \ptrue )$ is the pair $(
\psi, \emptyset )$, where after simplification, $\psi$ is the
$\extendsig{\Sigma}$-formula
%
\begin{eqnarray} \label{eq:ex-after}
\begin{split}
\forall y\typ{\fargsort{s}}.\;\, \lite( & \farg{s}( y ) \leq 0,  \\
 & s(\farg{s}( y )) \teq 0, \\
 & s(\farg{s}( y )) \teq \farg{s}( y ) + s( \farg{s}( y )-1 ) ) \\
 & \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq \farg{s}( y )-1 )
\end{split}  
\wedge ( s( c ) > 100 \wedge \exists z\typ{\fargsort{s}}. \farg{s}( z ) \teq c )
\end{eqnarray} 
%
The universal quantification in the resulting formula is over an uninterpreted
type $\fargsort{s}$, thus enabling finite model finding
techniques~\cite{ReyEtAl-1-RR-13} to search for a finite interpretation for
$\fargsort{s}$. Moreover, all occurrences of $y$ in this formula are beneath
applications of the uninterpreted function $\farg{s}$, implying that the
formula is in the essentially uninterpreted fragment of~\cite{GeDeM-CAV-09},
for which there exists a complete instantiation procedure. We observe that both
\cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}), and both
produce a model for formula~(\ref{eq:ex-after}) in $<0.1$ second. $\square$
\end{example}
